Iterative Monomorphisation

### 2024-02-15 iterative_monomorphisation.tex
started working on background

One of the main applications of automatic theorem provers is making the use of proof assistants easier. Many proof assistants support rank-1 polymorphism, however, many automatic theorem provers only operate on monomorphic logics. One approach to close this gap is extending existing provers, allowing them to natively support polymorphism, as was the case for Vampire \cite{vamp}. The main drawbacks of this solution are the amount of work required to extend these provers and the fact that it needs to be done from scratch for every monomorphic prover.

@@

The alternative would be to translate polymorphic problems into monomorphic problems. There are two main approaches to this, the first is a complete encoding of polymorphism using guards and tags \cite{mono-trans} into a monomorphic logic. The second solution is an incomplete encoding of polymorphism based on iterative monomorphisation, as introduced by B\"ome \cite[Section 2.2.1]{sb-phd}. This method relies on heuristically instantiating type variables with concrete types.

@@

  %* By a type version of compactness, if a set of polymorphic
  %  formulas is unsatisfiable, then there exists a finite set of monomorphic
  %  instances of the formulas that is also unsatisfiable. However, this
  %  finite set cannot be computed \cite[Section 2, Theorem 1]{expr-poly-types}; hence any approach based on
  %  instantiation of type variables is incomplete.

B\"ome's iterative approach is implemented as part of the SMT integration \cite[Chapter 2]{sb-phd} in Isabelle/HOL. This implementation is also used by Sledgehammer \cite{judgement, hammer} to interface with superposition based automatic theorem provers. However, it is is only briefly documented as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.

@@

In this paper, we present an algorithm based on his description and implementation (Section~\ref{sec:high-level-algorithm}). We also provide a more detailed description to allow the algorithm to be implemented by following the paper. Additionally, this descriptions shows some of the ways in which an implementation can be made more efficient and avoid explosions (Section~\ref{sec:low-level-algorithm}).

@@

The algorithm works as follows. We assume problems to be sets of formulas. All type arguments of the problem are collected, and the polymorphic type arguments are matched against the monomorphic ones. This yields new type arguments, both polymorphic and monomorphic. The process is then repeated using the newly derived type arguments.

@@

To keep the number of generated formulas finite, we limit the number of iterations. After the iterations are completed, the new type arguments are used to instantiate the type variables in the problem's formulas, generating new monomorphic clauses. Finally, because monomorphic provers generally do not support $n$-ary type constructors, types must be mangled. For example, the compound type $\ty{list}(\ty{int})$ could be mangled to $\ty{list\_int}$.

@@

% mention alternatives, (guards, tags) and current implementation in Isabelle

Language: TF1  %TH1?

We want our starting logic to be:
\begin{itemize}
    \item higher order
    \item polymorphic: we want atomic types, type constructors and universally quantified type variables
\end{itemize}
our target logic to be the same thing except we forego type constructors and type variables 
we want our target logic to be a subset of the starting logic

Definitions :
\begin{itemize}
   \item type arguments % part of the logic
   \item monomorphic and polymorphic types
   \item matching 
   \item term, literal, clause  % also part logic
\end{itemize}

@@

    It is important to paramatrise \(U\) with the clause of \(f\) because type variables are quantified at the clause level. This is not necessary for \(I\) because its type arguments do not contain type variables. [seems a sufficient explanation, if not clear expand further]

@@

[Add 2 or 3 examples to show how it works and explain why we have the different steps]

\subsection{Soundness}
Soundness of the algorithm follows from the fact that the resulting clauses are obtained by directly instantiating universally quantified type variable of the initial clauses.

Then we rely upon the soundness of mangling [encoding mono and poly ty].

\subsection{Incompleteness}

This algorithm is incomplete, this is because not all possible instantiations of the type variables are generated.
Consider the following problem [note that we need a consistent notation for type arguments (notably to distinguish them from normal arguments) also this might not be an issue but they are terms not literals ]: 
\begin{align*}
   P = \{ & \forall \alpha. f \langle \alpha \rightarrow int \rangle \\
          & \forall \alpha. f \langle nat \rightarrow \alpha \rangle \\
          & \forall \alpha \beta. \lnot f \langle \alpha \rightarrow \beta \rangle \}
\end{align*}

The algorithm only matches against ground type arguments. This is the case because polymorphic type unification is undecidable (quick ref) and whilst interweaving phases of unification with the current algorithm would be theoretically possible, it would most likely not be of any practical use. Indeed, the current implementation is already very explosive and generating new substitutions with a costly unification procedure would most likely not result in any improvements.

The previous problem would therefore never instantiate a clause, despite the problem becoming trivially contradictory if the last clause and any of the first two are instantiated.

Furthermore, the finite number of iterations of the algorithm constitutes an additional source of incompleteness in practice.
%find example that would require a list list list int to be solved but only allow such a type to be generated after 3 iterations

>>>>

One of the main applications of automatic theorem provers is reducing the workload when using proof assistants. Many proof assistants support rank-1 polymorphism, however, many automatic theorem provers only operate on monomorphic logics. One approach to close this gap is extending existing provers, allowing them to natively support polymorphism, as was the case for Vampire \cite{vamp}. The main drawbacks of this solution are the amount of work required to extend such provers and that it needs to be done from scratch for every monomorphic prover.

@@

The alternative would be to translate polymorphic problems into monomorphic problems. There are two main ideas to do this, the first is a complete encoding of polymorphism using guards or tags \cite{mono-trans} into a monomorphic logic. The second solution is an incomplete encoding of polymorphism based on iterative monomorphisation, as introduced by B\"ome \cite[Section 2.2.1]{sb-phd}. This method relies on heuristically instantiating type variables with concrete types.

@@

  %* By a type version of compactness, if a set of polymorphic
  %  formulae is unsatisfiable, then there exists a finite set of monomorphic
  %  instances of the formulae that is also unsatisfiable. However, this
  %  finite set cannot be computed \cite[Section 2, Theorem 1]{expr-poly-types}; hence any approach based on
  %  instantiation of type variables is incomplete.

B\"ome's iterative approach is implemented as part of the SMT integration \cite[Chapter 2]{sb-phd} in Isabelle/HOL. This implementation is also used by Sledgehammer \cite{judgement, hammer} to interface with superposition based automatic theorem provers. However, it is is only documented as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.

@@

In this paper, we present an algorithm based on his description and implementation (Section~\ref{sec:high-level-algorithm}). We also provide a more detailed description to allow the algorithm to be implemented by following the paper. Additionally, this description shows some of the ways in which an implementation can be made more efficient and avoid explosions (Section~\ref{sec:low-level-algorithm}).

@@

The algorithm works as follows. We assume problems to be sets of formulae. All type arguments of the problem are collected, and the polymorphic type arguments are matched against the monomorphic ones. This yields new type arguments, both polymorphic and monomorphic. The process is then repeated using the newly derived type arguments.

@@

To keep the number of generated formulae finite, we limit the number of iterations. After the iterations are completed, the new type arguments are used to instantiate the type variables in the problem's formulae, generating new monomorphic clauses. Finally, because monomorphic provers generally do not support $n$-ary type constructors, types must be mangled. For example, the compound type $\ty{list}(\ty{int})$ could be mangled to the constant $\ty{list\_int}$.

@@

% mention alternatives, (guards, tags) and current implementation in Isabelle
Definitions :
\begin{itemize}
   \item our logic (TF1), our starting and target logics (with and without polymorphism)
   \item notations
   \item type arguments
   \item monomorphic and polymorphic types
   \item matching 
   \item term, literal, clause 
\end{itemize}

% TH1 ugly for height of 1 reasons, can be fixed with case feature
Our starting logic will be TPTP's TH1 \cite{th1}, a higher-order logic with rank-1 polymorphism. We will assume the initial problem to be TH1 formulae.

types are either constants, type variables, or a type constructor of a certain arity applied to other types

we consider a set of function symbols, with type arguments and arguments of a certain arity

a problem is made up of a set of formulae, type variables are quantified at the formula level

note that in the implementation, a problem is a conjunction of disjunctive clauses, this is only relevant when the problem is parsed when the initial type arguments are collected and at the very end when the final clauses are generated

@@

    It is important to paramatrise \(U\) with the clause of \(f\) because type variables are quantified at the clause level. This is not necessary for \(I\) because its type arguments do not contain type variables.

@@

This algorithm is clearly sound as the newly derived clauses are instances of the initial clauses where universally quantified types have been replaced by concrete types.
